### 2022-10-26 manuscript/manuscript.tex
fixes

@@

In their 2016 paper, Santiago and Caballero derive a general equation for the
fitness variance that accounts for the reduction due to the fixation of weakly
deleterious alleles (i.e. the rate of Muller's ratchet). There derivation for
the variance uses a heuristic approach based on Fisher's Fundamental Theorem of
Natural Selection (equation 2, \cite{Santiago2016-mu}, equation 15A
\cite{Garcia-Dorado2007-jj}). A more formal derivation follows from
\textcite{Higgs1995-xc} (though they assume an asexual system, the result is
the same). In their work, they derive equations for the change per generation
for the moments and cross-moments of the fitness distribution under
multiplicative selection in a haploid population. For a Poisson number of new
mutations per gamete, per generation, they find (equation 13.3,
\cite{Higgs1995-xc}) that the rate of the ratchet is

\begin{align}
  R = U - s \V(n).
\end{align}

The rate of the ratchet for a diploid population is the same, as long as $U$ is
the mutation rate per generation per diploid genome (as is assumed here).
$\V(n)$ is the variance of the distribution of the \emph{number} of deleterious
mutations (not the fitness distribution); as long as selection is not too
strong, $V_A \approx -s^2\V(n)$. Then, the rate of the ratchet in terms of the
fitness variance is,

\begin{align}
  %R &= U - s\V(n) \\
  R &\approx U + \frac{V_A}{s} \\
  V_A &\approx Us - Rs. 
\end{align}

This is identical to their equation 2, noting that the waiting time between
fixations (i.e. ratchet clicks) is $T = \nicefrac{1}{R}$. Note that if selected
sites are sufficiently deleterious that they cannot fix (or the population is
infinite), $R=0$ and the additive fitness variance is identical to that under
strong background selection, $V_A = Us = V_\text{BGS}$ (equation
\eqref{eq:va_bgs}).

Then, the fixation probability of a new deleterious mutation with additive
effects ($1-s$ in the heterozygote, $1-2s$ in the homozygote) in a diploid
population is,

@@

(c.f. \cite{Durrett2008-ql}, equation 7.21; \cite{Kimura1957-rk}, equation
5.6). Then, Santiago and Caballero argue that the rate of the ratchet is the
inverse of the averaging waiting time until a fixation, $T \approx 1/(2UN
p_\text{fix})$, so

\begin{align}
  T \approx \frac{\exp(4 N_e s)-1}{4 U s N_e}.
\end{align}

@@

\begin{align}
  R(N_e) := \frac{4 U s N_e}{\exp(4 N_e s)-1}
\end{align}

@@

\begin{align}
  R(N_{e,\infty}) &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
  N_{e,\infty} &= N \exp\left(-\frac{(U-R(N_{e,\infty}))s}{\left(1-\frac{Us}{U-R(N_{e,\infty})}\right) \left(2-(2-L) \frac{Us}{U-R(N_{e,\infty})}\right)} \right).
\end{align}

>>>>

TODO we need to make appoint about how Ben G's paper found the V(n)->VA connectoin -- this is their patching model

@@

% NOTE: The S&C math using Fisher's FToNS argues fitness decreases
% at a rate of -s/T each generation. However their model is haploid;
% in a diploid model, fixation makes everyone  a homozygote with fitness
% effect -2s; thus the rate must be -2s/T which agrees with sim

In their 2016 paper, Santiago and Caballero derive a general equation for the
fitness variance that accounts for the reduction due to the fixation of weakly
deleterious alleles (i.e. the rate of Muller's ratchet). There derivation for
the variance uses a heuristic approach based on Fisher's Fundamental Theorem of
Natural Selection (equation 2, \cite{Santiago2016-mu}, equation 15A
\cite{Garcia-Dorado2007-jj}). A more formal derivation follows from
\textcite{Higgs1995-xc} (though they assume a haploid asexual system, the end
result is the same up to a factor of two). In their work, they derive equations
for the change per generation for the moments and cross-moments of the fitness
distribution under multiplicative selection in a haploid model. They find, the
haploid rate of the ratchet is $r = u - s v_n$ (equation 13.3,
\cite{Higgs1995-xc}), where we use lowercase to distinguish between haploid and
diploid models. Here, $v_n$ is the variance in the \emph{number} of deleterious
mutations; as long as selection is not too strong, the haploid additive
variance is $v_{A} \approx s^2 v_n$. Substituting this for $v_n$ and
rearranging, the haploid fitness variance is related to the $v_A = us - rs$,
which is identical to Santiago and Caballero's (\citeyear{Santiago2016-mu})
equation 2, noting that the waiting time between fixations (i.e. ratchet
clicks) is $T = \nicefrac{1}{r}$. Intuitively, the fitness variance $Us$ is
reduced by the rate at which deleterious mutations fix and remove fitness
variation from the population.

Now, note that under a diploid model, the fitness variance is twice that as the
haploid model, so $V_A = 2 v_A$. Then, the fitness variance is,

\begin{align}
  V_{A} = Us - 2Rs. 
\end{align}

where $U = 2u$ and $R$ is the diploid rate of the ratchet, which depends on
higher moments of the fitness distribution, but only differs from $r$ through a
rescaling of the population size. Note that if selected sites are sufficiently
deleterious that they cannot fix (or the population is infinite), $R=0$ and the
additive fitness variance is identical to that under strong background
selection, $V_A = Us = V_\text{BGS}$ (equation \eqref{eq:va_bgs}). Then, the
fixation probability of a new deleterious mutation with additive effects ($1-s$
in the heterozygote, $1-2s$ in the homozygote) in a diploid population is,

@@

(c.f. \cite{Durrett2008-ql}, equation 7.21; \cite{Kimura1957-rk}, equation
5.6). Then, Santiago and Caballero argue that the rate of the ratchet is the
inverse of the averaging waiting time until a fixation, $T \approx 1/(UN
p_\text{fix})$ (note that since $U$ is the \emph{diploid} mutation rate, the
total population mutation rate is $UN$ per generation), so

\begin{align}
  T \approx \frac{\exp(4 N_e s)-1}{2 U s N_e}.
\end{align}

@@

\begin{align}
  R(N_e) := \frac{2 U s N_e}{\exp(4 N_e s)-1}
\end{align}

@@

\begin{align}
  R(N_{e,\infty}) &= \frac{4 U s N_{e,\infty}}{\exp(4 N_{e,\infty} s)-1}\\
  N_{e,\infty} &\approx N \exp\left(-\frac{V_A}{(1-Z) (2-(2-L) Z)} \right) \\
  N_{e,\infty} &= N \exp\left(-\frac{(U-R(N_{e,\infty}))s}{\left(1-\frac{Us}{U-R(N_{e,\infty})}\right) \left(2-(2-L) \frac{Us}{U-R(N_{e,\infty})}\right)} \right).
\end{align}
